Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-29 paper/main.tex
use nameref instead of ref

By utilizing the provided data extraction methods, we can infer properties of the bacteria that aid
us in constructing a suitable computational model.
However, the construction of any mechanistic model also requires prior konwledge which can be used
to inform our analysis methods.
For example, by extracing the positions of bacteria across a time-series of microscopic images, we
can also calculate their lengths at these time-points and thus determine their rate of growth and
type of growth (see section~\ref{paragraph:parameter-estimation-individual-treatment}), thus
effectively guiding our modeling steps.
As we will show later, extracted data in the form of cellular positions or segmentation masks can
also be used for comparing computational results with biological ground truth.

@@

\begin{figure}[H]
    \centering
    \includestandalone[width=0.8\textwidth]{figures/concept/overview}
    \caption{
        The project can be divided into subcomponents that interact with and depend on
        each other.
        Every component implicitly builds on the spatial representation of rod-shaped bacteria as
        vertices (see section~\ref{subsec:spatial-representation}).
        Red boxes highlight the key scientific drivers of this study while yellow boxes indicate
        auxiliary methods.
        The blue box "Data Generation" displays a extension of this approach which is under active
        development.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

@@

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod into two new agents.\\
            % Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (4) & Adhesion &
            Bacteria attract at longer distances and adhere to each other at shorter
            distances~\cite{Verwey1947,Trejo2013}.\\
        \bottomrule
    \end{tabularx}
    \caption{
        This table lists aspects of cellular behaviour which we consider in our computational model.
        It is split into cellular aspects (C) which concern only an individual bacterium and
        interactions between cells (CC).
        These aspects are generalized in order to fit to as many types of rod-shaped bacteria as
        possible.
        These are the assumptions for the basic model that we are presenting here.
        In order for our methods to work, we are really only required to satisfy the assumption of
        our spatial representation (see section~\ref{subsec:spatial-representation}) and thus, these
        assumptions can also be loosened if required.
    }
    \label{table:simulation-aspects}
\end{table}

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
By varying its parameters, the Mie potential is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels that correspond to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation
(see section~\ref{sec:supplement-morse-potential}).
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Numerical Implementation}
We implemented the described model in our recently developed \ac{abm} framework
\texttt{cellular\_raza}~\cite{Pleyer2025}.
By design \texttt{cellular\_raza} does not assume any particular cellular representation, which
makes it flexible and easy to adapt the simulation to particular needs.
Its design allows us to experiment with different model implementations and facilitates the
explorative capabilities of researchers.
It is furthermore able to seamlessly integrate into Python code-bases thus allowing us to utilize
methods from other popular software libraries such as \texttt{numpy} and \texttt{scipy}.
Furthermore, it is an \ac{abm} which is able to natively track mother-daughter relations of cells
over the course of a numerical simulation which is an integral component of our image-space metric
comparison algorithm (see section~\ref{subsec:image-space-metric}).
Further details can be found in the supplement (see section~\ref{section:supplement-performance})

%###################################################################################################
\section{Methodology}
\label{sec:methodology}
These sections introduce methods which can be combined in order to estimate the parameters of our
computational model.
It is split into 3 parts, each covering distinct purposes.
We begin by explaining the visualization process which will later be used within our image-space
metric (see section~\ref{subsec:image-space-metric}) and for testing purposes.
Afterwards, we introduce a position extraction algorithm which extracts cellular information in the
form of polygons from segmented microscopic images.
Finally, we will provide two distinct comparison algorithms which can be used to compare numerically
obtained results with experimental data.
We will test and benchmark the constructed algorithms within the respective sections.
As highlighted before, the developed methods only depend on the spatial representation described in
section~\ref{subsec:spatial-representation}.
Consequently, these methods can be applied to a broader class of mechanical models that can exhibit
varied cellular behaviours.

@@

This section explains how we can transform a segmented microscopic image such that we obtain a
collection of vertices (see Section~\ref{subsection:mechanical-model-mechanics}) that can be used
to initialize our agents and to perform parameter estimation.
There exists a plethora of tools to perform segmentation of
images~\cite{Stringer2020,Cutler2022,Hardo2022}
Figure~\ref{fig:position-extraction-algorithm} (A) shows a typical microscopic image.
Our algorithm requires that images are prepared using a pixel-based (in-image) encoding scheme, as
we explained in section~\ref{subsection:visualization}.

@@

In order to quantitatively analyze this disparity we use both data to calculate the average rod
length of the bacteria (Figure~\ref{fig:benchmarking-extraction-algorithm} (J)) over the course of
the simulated time interval.
The results show good agreement with larger discrepancies directly after division events.
In addition figure~\ref{fig:benchmarking-extraction-algorithm} (K) displays individual displacements
of exact and extracted vertices which have been reoriented with respect to the direction of the rod.
Values closer to the center represent better agreement.
The distribution of these displacements can be well described by a multivariate normal
distribution with mean $\mu$ and variances $\sigma_{ij}$.
These parameters can be obtained by fitting the distribution to the displacements at every time
point.
The parameters resulting from this fit are displayed in
Figure~\ref{fig:benchmarking-extraction-algorithm} (L), where the overall variance (filled area) in
the displacements significantly increases directly after division events.
However, the centers of the distributions (solid curves) remain concentrated near zero without
significant outliers which demonstrates the robustness of our method in these densely packed
scenarios.
These observations provide us with enough information to construct an error model for the extraction
algorithm.
Further details such as information about the covariances are given in the supplementary materials
(section~\ref{sec:supplement-extraction-algorithm}).

@@

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/data/crm_fit/001044.png}%
        \node at (0.02, 0.98)[anchor=north west, white]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/data/crm_fit/001044-markers.pdf}%
        \node at (0.02, 0.98)[anchor=north west, white]{\textbf{B}};
    \end{tikzonimage}\\
    \vspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.28\textwidth]
        {figures/crm_fit-data-progression.png}
        \node at (0.02, 0.98)[anchor=north west, white]{\textbf{C}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.35\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-individual.pdf}%
        \node at (0, 1)[anchor=north west, black]{\textbf{D}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.35\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/growth-rates-bar-plot.pdf}%
        \node at (0, 1)[anchor=north west, black]{\textbf{E}};
    \end{tikzonimage}%
    \caption{
        \textbf{TODO also insert vertical in (B) line to indicate when direct comparison algorithm
        starts.}
        \textbf{Link to supplement for discussion of bar plot}\\
        (A) Snapshot of used microscopic images.
        (B) Segmented microscopic image.
        (C) Selected time points of the utilized time series showcasing the spatial configurations
        of the investigated bacteria.
        A complete list of snapshots is included in the supplementary material (see
        section~\ref{sec:supplement-data-details}).
        (D) Growth curves of individual bacteria including a discrete delay before exponential
        growth is experienced.
        (E) Distribution of growth parameters shows cell-cell variability of the analyzed sample.
        The obtained results are further discussed in the supplementary material section~\ref{}.
    }
    \label{fig:application-2-data-overview}
\end{figure}

@@

The image-space metric is computationally much more expensive compared to the direct method.
There are 4 main steps which are executed within any single optimization iteration that contribute
to the total wall time.
These steps are prediction via the computational model ($15.63(21)\%$), mask adjustment of the data
masks ($14.49(19)\%$), mask rendering for the numerical results ($69.2(1.4)\%$) and finally
comparison of the masks ($0.68(5)\%$) (see section \ref{section:supplement-performance}).
The second step alters the masks provided by the data for every individual numerical simulation.
This is required since we need to ensure that the colors of the data-masks match the colors of the
masks produced by our computational model.
To this end, we need to construct a family tree of the data-bacteria and compare with the
\textit{in-silico} family tree.
We can see that the generation of masks from the numerical results takes up the majority of the work.
In addition, the step of mask adjustment would not have been required within the position-comparison
algorithm.
Combining these two values, the position-comparison algorithm is $\approx 6.13(8)$ times faster.

>>>>

By utilizing the provided data extraction methods, we can infer properties of the bacteria that aid
us in constructing a suitable computational model.
However, the construction of any mechanistic model also requires prior konwledge which can be used
to inform our analysis methods.
For example, by extracing the positions of bacteria across a time-series of microscopic images, we
can also calculate their lengths at these time-points and thus determine their rate of growth and
type of growth (see \nameref{paragraph:parameter-estimation-individual-treatment}), thus
effectively guiding our modeling steps.
As we will show later, extracted data in the form of cellular positions or segmentation masks can
also be used for comparing computational results with biological ground truth.

@@

\begin{figure}[H]
    \centering
    \includestandalone[width=0.8\textwidth]{figures/concept/overview}
    \caption{
        The project can be divided into subcomponents that interact with and depend on
        each other.
        Every component implicitly builds on the spatial representation of rod-shaped bacteria as
        vertices (see \nameref{subsec:spatial-representation}).
        Red boxes highlight the key scientific drivers of this study while yellow boxes indicate
        auxiliary methods.
        The blue box "Data Generation" displays a extension of this approach which is under active
        development.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

@@

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabularx}{\textwidth}{c l X}
        &\textbf{Aspect} & \textbf{Description}\\
        \toprule
        &\textbf{(C) Cellular}\\
        \midrule
        (1) & Rod-Shaped Mechanics &
            Rod-shaped bacteria are flexible rods which are able to freely move around (off-lattice
            approach) \cite{Takeuchi2005,Ursell2014,Amir2014_2}.\\
        (2) & Growth &
            Cells grow exponentially by inserting new material either along the circular part of the
            rod or at the tip~\cite{Robert2014,Takeuchi2005}.\\
        (3) & Division &
            Rod-Shaped bacteria divide in the middle of the rod into two new agents.\\
            % Bacteria can form multilayers during their growth phase \cite{Duvernoy2018}.\\
        &\textbf{(CC) Cell-Cell Interactions}\\
        \midrule
        (4) & Adhesion &
            Bacteria attract at longer distances and adhere to each other at shorter
            distances~\cite{Verwey1947,Trejo2013}.\\
        \bottomrule
    \end{tabularx}
    \caption{
        This table lists aspects of cellular behaviour which we consider in our computational model.
        It is split into cellular aspects (C) which concern only an individual bacterium and
        interactions between cells (CC).
        These aspects are generalized in order to fit to as many types of rod-shaped bacteria as
        possible.
        These are the assumptions for the basic model that we are presenting here.
        In order for our methods to work, we are really only required to satisfy the assumption of
        our spatial representation (see \nameref{subsec:spatial-representation}) and thus, these
        assumptions can also be loosened if required.
    }
    \label{table:simulation-aspects}
\end{table}

@@

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
By varying its parameters, the Mie potential is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels that correspond to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation
(see \nameref{sec:supplement-morse-potential}).
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Numerical Implementation}
We implemented the described model in our recently developed \ac{abm} framework
\texttt{cellular\_raza}~\cite{Pleyer2025}.
By design \texttt{cellular\_raza} does not assume any particular cellular representation, which
makes it flexible and easy to adapt the simulation to particular needs.
Its design allows us to experiment with different model implementations and facilitates the
explorative capabilities of researchers.
It is furthermore able to seamlessly integrate into Python code-bases thus allowing us to utilize
methods from other popular software libraries such as \texttt{numpy} and \texttt{scipy}.
Furthermore, it is an \ac{abm} which is able to natively track mother-daughter relations of cells
over the course of a numerical simulation which is an integral component of our image-space metric
comparison algorithm (see \nameref{subsec:image-space-metric}).
Further details can be found in the supplement (see \nameref{section:supplement-performance})

%###################################################################################################
\section{Methodology}
\label{sec:methodology}
These sections introduce methods which can be combined in order to estimate the parameters of our
computational model.
It is split into 3 parts, each covering distinct purposes.
We begin by explaining the visualization process which will later be used within our image-space
metric (see \nameref{subsec:image-space-metric}) and for testing purposes.
Afterwards, we introduce a position extraction algorithm which extracts cellular information in the
form of polygons from segmented microscopic images.
Finally, we will provide two distinct comparison algorithms which can be used to compare numerically
obtained results with experimental data.
We will test and benchmark the constructed algorithms within the respective sections.
As highlighted before, the developed methods only depend on the spatial representation described in
section~\ref{subsec:spatial-representation}.
Consequently, these methods can be applied to a broader class of mechanical models that can exhibit
varied cellular behaviours.

@@

This section explains how we can transform a segmented microscopic image such that we obtain a
collection of vertices (see \nameref{subsection:mechanical-model-mechanics}) that can be used
to initialize our agents and to perform parameter estimation.
There exists a plethora of tools to perform segmentation of
images~\cite{Stringer2020,Cutler2022,Hardo2022}
Figure~\ref{fig:position-extraction-algorithm} (A) shows a typical microscopic image.
Our algorithm requires that images are prepared using a pixel-based (in-image) encoding scheme, as
we explained in section~\ref{subsection:visualization}.

@@

In order to quantitatively analyze this disparity we use both data to calculate the average rod
length of the bacteria (Figure~\ref{fig:benchmarking-extraction-algorithm} (J)) over the course of
the simulated time interval.
The results show good agreement with larger discrepancies directly after division events.
In addition figure~\ref{fig:benchmarking-extraction-algorithm} (K) displays individual displacements
of exact and extracted vertices which have been reoriented with respect to the direction of the rod.
Values closer to the center represent better agreement.
The distribution of these displacements can be well described by a multivariate normal
distribution with mean $\mu$ and variances $\sigma_{ij}$.
These parameters can be obtained by fitting the distribution to the displacements at every time
point.
The parameters resulting from this fit are displayed in
Figure~\ref{fig:benchmarking-extraction-algorithm} (L), where the overall variance (filled area) in
the displacements significantly increases directly after division events.
However, the centers of the distributions (solid curves) remain concentrated near zero without
significant outliers which demonstrates the robustness of our method in these densely packed
scenarios.
These observations provide us with enough information to construct an error model for the extraction
algorithm.
Further details such as information about the covariances are given in the supplementary materials
(see \nameref{sec:supplement-extraction-algorithm}).

@@

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/data/crm_fit/001044.png}%
        \node at (0.02, 0.98)[anchor=north west, white]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.49\textwidth]
        {figures/data/crm_fit/001044-markers.pdf}%
        \node at (0.02, 0.98)[anchor=north west, white]{\textbf{B}};
    \end{tikzonimage}\\
    \vspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.28\textwidth]
        {figures/crm_fit-data-progression.png}
        \node at (0.02, 0.98)[anchor=north west, white]{\textbf{C}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.35\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-individual.pdf}%
        \node at (0, 1)[anchor=north west, black]{\textbf{D}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.35\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/growth-rates-bar-plot.pdf}%
        \node at (0, 1)[anchor=north west, black]{\textbf{E}};
    \end{tikzonimage}%
    \caption{
        \textbf{TODO also insert vertical in (B) line to indicate when direct comparison algorithm
        starts.}
        \textbf{Link to supplement for discussion of bar plot}\\
        (A) Snapshot of used microscopic images.
        (B) Segmented microscopic image.
        (C) Selected time points of the utilized time series showcasing the spatial configurations
        of the investigated bacteria.
        A complete list of snapshots is included in the supplementary material (see
        \nameref{sec:supplement-data-details}).
        (D) Growth curves of individual bacteria including a discrete delay before exponential
        growth is experienced.
        (E) Distribution of growth parameters shows cell-cell variability of the analyzed sample.
        The obtained results are further discussed in the supplementary material section~\ref{}.
    }
    \label{fig:application-2-data-overview}
\end{figure}

@@

The image-space metric is computationally much more expensive compared to the direct method.
There are 4 main steps which are executed within any single optimization iteration that contribute
to the total wall time.
These steps are prediction via the computational model ($15.63(21)\%$), mask adjustment of the data
masks ($14.49(19)\%$), mask rendering for the numerical results ($69.2(1.4)\%$) and finally
comparison of the masks ($0.68(5)\%$) (see section \nameref{section:supplement-performance}).
The second step alters the masks provided by the data for every individual numerical simulation.
This is required since we need to ensure that the colors of the data-masks match the colors of the
masks produced by our computational model.
To this end, we need to construct a family tree of the data-bacteria and compare with the
\textit{in-silico} family tree.
We can see that the generation of masks from the numerical results takes up the majority of the work.
In addition, the step of mask adjustment would not have been required within the position-comparison
algorithm.
Combining these two values, the position-comparison algorithm is $\approx 6.13(8)$ times faster.
